---
title: 'Assignment 2: Data visualization'
author: "Peter Szecsi"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(scales)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
#Filter Data#
exp <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/expeditions.csv")
exp <- exp %>%
  group_by(peak_name) %>%
  mutate(n_exp = n())

temp <- exp[,c("peak_name", "n_exp")]
temp <- temp %>%
  unique() %>%
  arrange(n_exp)

temp <-  filter(temp, n_exp >= temp[(nrow(temp)-14), 2])

exp <- exp %>%
  filter(peak_name %in% temp$peak_name) %>%
  arrange(desc(n_exp))

#plotting#

plot1 <- ggplot(data = exp, aes(reorder(peak_name, n_exp))) +
  geom_bar(aes(fill = season)) +
  scale_fill_viridis_d() +
  coord_flip() +
  theme_bw(base_size = 15.8) +
  theme(legend.position = "bottom") +
  labs(title="The 15 most popular peaks stacked by season expedition",
        x ="", y = "Number of expeditions")
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
library(RColorBrewer)

phd <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-19/phd_by_field.csv")

phd <- phd %>%
  group_by(broad_field, year) %>%
  mutate(awards = sum(n_phds, na.rm = TRUE)) %>%
  ungroup()


plot2 <- ggplot(data = phd, aes(year, awards, group = broad_field, colour = broad_field)) +
  geom_line(size = 1.2) +
  labs(colour = "Broad field", title = "Number of awarded Ph.D.-s in the US by year",
      x = "", y = "") +
  theme_minimal(base_size = 15.5) +
  scale_x_continuous(breaks = pretty_breaks(5)) +
  scale_colour_brewer(palette = "Dark2") +
  theme(panel.border = element_blank()) +
  scale_y_continuous(labels = comma_format())
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
bkv <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-05/commute.csv")

bkv <- bkv %>%
  group_by(state, mode) %>%
  mutate(sum_n = sum(n, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(log_sum_n = log(sum_n))

tempW <- filter(bkv, mode == "Walk")[,c("state", "state_region", "sum_n", "log_sum_n", "state_abb")] %>%
  rename(w_sum_n = sum_n) %>%
  rename(w_log_sum_n = log_sum_n) %>%
  unique()

tempW <- tempW[complete.cases(tempW),]

tempB <- filter(bkv, mode == "Bike")[,c("state", "state_region", "sum_n", "log_sum_n")] %>%
  rename(b_sum_n = sum_n) %>%
  rename(b_log_sum_n = log_sum_n) %>%
  unique()

tempB <- tempB[complete.cases(tempB),]

dat <- left_join(tempW, tempB, by = c("state", "state_region"))


plot3 <- ggplot(data = dat, mapping = aes(x = w_log_sum_n, y = b_log_sum_n)) +
  geom_point(aes(colour = factor(state.region))) +
  geom_text(aes(label = state_abb), check_overlap = TRUE) +
  theme_light() +
  scale_y_continuous(labels = comma_format()) +
  scale_x_continuous(labels = comma_format()) +
  labs(title="Title number of people walking vs. biking to work in each USA state",
      x ="Number of ppl walking to work (log N)", y = "Number of ppl biking to work (log N)",
      colour = "State Region")
plot3

```
